Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-04-13 main.Rmd
quasi-circular section: diagram and examples update

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\hline
quasi-circular granularity && potential period lengths\\
\hline
$Q_1 =$ day-of-month && $P_1 = 31, 30, 29, 28$\\
$Q_2 =$ hour-of-month && $P_2 = 24*31, 24*30, 24*29, 24*28$\\
$Q_3 =$ day-of-year && $P_3 = 366, 365$\\
$Q_4 =$ week-of-month && $P_4 = 5, 4$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of quasi-circular granularities $Q_i$ with potential period lengths $P_i$. These quasi-circular granularities relate two linear granularities one of which groups periodically into the other with irregular mapping and hence can not be expressed through modular arithmetic. }
\label{tab:quasi}
\end{table}

\begin{definition}\label{def:quasicircular}
A {\bf quasi-circular granularity} $Q_{B, G'}$ relates linear granularities $G'$ and bottom granularity $B$, if
\begin{equation}\label{eq:quasi}
Q_{B, G'}(z) = z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert,for\quad z \in T_k
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$B$ denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G'$ with irregular mapping, $P'$ and $R'$ are the period of the grouping $(B, G')$ and the number of granules of $G'$ in each of these periods, $T_w$ are the sets of indices of $B$ describing $G'(w)$ such that  $G'(w) = \bigcup_{z \in T_w}B(z)$ and $\vert T_w \vert$ is the cardinality of set $T_w$.
\end{definition}

@@

**Example**: 
<!-- Example showing quasi-circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:quasi-circular-example}. Each granule is represented by a box.  -->

Consider two linear granularities $G' = Alter^{2}_{(1,-1)}(BG, Group_3(B))$ and $H' = Alter^{2}_{(1,-2)}(BG, Group_7(B))$.<!--This implies that $G'$ is made up by shrinking every 1st granule of $Group_3(B)$ by $1$ granule and $H'$ is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules. Number of granules of $G'$ and $H'$ in each period of $B$ is 2 but the number of granules of $B$ in each of those granules are different.--> Then $Q_{B, G'}$ and $Q_{B, H'}$ are repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not necessarily the same across different granules of $G'$ or $H'$. For $G'$, $T_0 = \{0,1\}$ and $T_1 = \{2, 3, 4\}$. For $H'$, $T_0 = \{0, 1, 2, 3, 4, 5, 6\}$ and $T_1 = \{7, 8, 9, 10, 11\}$, then we will have \autoref{eq:eq4} and  \autoref{eq:eq5}.

\begin{equation} \label{eq:eq4}
\begin{split}
Q_{B,G'}(8) & = L(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) ,\quad since \quad 8 \in T_{3} \\
  & = L(1) \\
\end{split}
\end{equation}

@@

\begin{equation} \label{eq:eq5}
\begin{split}
Q_{B,H'}(10) & = L(10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert) ,\quad since \quad  10 \in T_{1}  \\
  & = L(3) \\
\end{split}
\end{equation}

**Discussion**: If linear granularity $G'$ is periodical with respect to $B$ with irregular mapping, then there exist $R'$, $P' \in \mathbb{Z}_+$ such that if $G'(w) = \bigcup_{z \in T_w}B(z)$ then $$G'(w) = \bigcup_{z \in T_w \mod R'}B(P'*\lfloor w/R' \rfloor + z)$$ (from definition \ref{def:periodical})<!-- - $B \trianglelefteq G'$ --> <!-- and,  number of granules of $B$ in each granule of $G'$ is not a constant (irregular mapping) -->. Here $w \mod R'$ represents the index that must be shifted to obtain $G'(w)$. The idea here is if we know the composition of each of the granules of $G'$ in terms of granules of $B$ for one period, we can find the composition of any granule of $G'$ beyond a period since the "pattern" repeats itself along the time domain due to the periodic property. The periodic property also ensures that $\vert T_w \vert = \vert T_{w\mod R'} \vert$ since every $w^{\text{th}}$ and $(w+R')^{\text{th}}$ granule of $G'$ will have the same number of granules of $B$. The term $\sum_{w=0}^{k-1}\vert T_{w}\vert$ denotes the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$. Since $\vert T_w \vert = \vert T_{w\mod R'} \vert$, the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$ becomes $\sum_{w=0}^{k-1}\vert T_{w \mod R'}\vert$ in Definition \autoref{def:quasicircular}.

```{r quasi-circular-example, echo = FALSE, out.width = "100%",fig.cap="Terminology and intuition of quasi-circular granularities are shown. Panel a shows the time domain distributed as linear granularities $B$ (bottom granularity), $G$, $H'$ and quasi-circular granularities $Q(B, G)$ and $Q(B, H')$. Both $Q(B, G)$ and $Q(B, H')$ have repeating patterns unlike $G$ or $H'$ and the patterns are guided by the distribution of $B$ within $G$ and $H'$ and remains constant within a period shown by the hashed arrows. Panel b depicts the intuition through a measuring tape analogy where the folds are not of same length but has same length over a period."}

@@

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="Terminology and intuition of aperiodic cyclic granularities are shown. Panel a shows the time domain distributed as linear granularities Day, Semester, within semester breaks, in-session and aperiodic cyclic granularity Week-of-Semester. $A(B, M)$ consists of three categories 0, 1, 2 non-semester period, within-semester breaks and in-session semester period respectively. Panel b depicts the intuition through a measuring tape analogy which needs to be folded at different points every year since $M1$ and $M2$ will fall in different weeks."}
knitr::include_graphics("Figs/quasi-periodic-example.png")
```

<!-- To consider the exhaustive set of temporal regularities that might exist in the data, we can also define temporal granularities based on the hierarchical structure of a calendar. -->

>>>>

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\hline
quasi-circular granularity && potential period lengths\\
\hline
$Q_1 =$ day-of-month && $P_1 = 31, 30, 29, 28$\\
$Q_2 =$ hour-of-month && $P_2 = 24*31, 24*30, 24*29, 24*28$\\
$Q_3 =$ day-of-year && $P_3 = 366, 365$\\
$Q_4 =$ week-of-month && $P_4 = 5, 4$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Examples of quasi-circular granularities $Q_i$ with potential period lengths $P_i$. These quasi-circular granularities relate two linear granularities one of which groups periodically into the other with irregular mapping leading many potential period lengths and hence can not be expressed through modular arithmetic. }
\label{tab:quasi}
\end{table}

\begin{definition}\label{def:quasicircular}
A {\bf quasi-circular granularity} $Q_{B, G'}$ relates linear granularities $G'$ and bottom granularity $B$, if
\begin{equation}\label{eq:quasi}
Q_{B, G'}(z) = z - \sum_{w=0}^{k-1}\vert T_{w \mod R'} \vert,for\quad z \in T_k
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$B$ denotes a full-integer labelled bottom granularity which groups periodically into linear granularity $G'$ with irregular mapping, $P'$ and $R'$ are the period of the grouping $(B, G')$ and the number of granules of $G'$ in $P'$, $T_w$ are the sets of indices of $B$ describing $G'(w)$ such that  $G'(w) = \bigcup_{z \in T_w}B(z)$ and $\vert T_w \vert$ is the cardinality of set $T_w$.
\end{definition}

@@

<!-- **Example**:  -->
<!-- Example showing quasi-circular granularities relating two linear granularities each with bottom granularities are visually depicted in a series of slots in \autoref{fig:quasi-circular-example}. Each granule is represented by a box.  -->

Consider a linear granularity $G'$ such that $G' = Alter^{2}_{(1,-2)}(B, Group_7(B))$, which implies it is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules.<!-- $$G'= Alter^{2}_{(1,-1)}(BG, Group_3(B))$$ and --><!--This implies that $G'$ is made up by shrinking every 1st granule of $Group_3(B)$ by $1$ granule and $H'$ is made up of shrinking every 1st granule of $Group_3(B)$ by $2$ granules. Number of granules of $G'$ and $H'$ in each period of $B$ is 2 but the number of granules of $B$ in each of those granules are different.--> Then $Q_{B, G'}$ is a repetitive categorization of time, similar to circular granularities, except that the number of granules of $B$ is not necessarily the same across different granules of $G'$. Here, $T_0 = \{0, 1, 2, 3, 4, 5, 6\}$ and $T_1 = \{7, 8, 9, 10, 11\}$. Hence using \ref:{def:quasicircular}, then we will have:

<!-- \begin{equation} \label{eq:eq4} -->
<!-- \begin{split} -->
<!-- Q_{B,G'}(8) & = L(8 - \sum_{w=0}^{3-1}\vert T_{w \mod 2}\vert) ,\quad since \quad 8 \in T_{3} \\ -->
<!--   & = L(1) \\ -->
<!-- \end{split} -->
<!-- \end{equation} -->

@@

\begin{align}
\begin{split}
Q_{B,H'}(10) & = 10 - \sum_{w=0}^{1-1}\vert T_{w \mod 2}\vert ,\quad since \quad  10 \in T_{1}  \\
  & = 3 \\
\end{split}
\end{align}

<!-- **Discussion**:  -->
If linear granularity $G'$ is periodical with respect to $B$ with irregular mapping, then there exist $R'$, $P' \in \mathbb{Z}_+$ such that if $G'(w) = \bigcup_{z \in T_w}B(z)$ then $$G'(w) = \bigcup_{z \in T_w \mod R'}B(P'*\lfloor w/R' \rfloor + z)$$ (from definition \ref{def:periodical})<!-- - $B \trianglelefteq G'$ --> <!-- and,  number of granules of $B$ in each granule of $G'$ is not a constant (irregular mapping) -->. Here $w \mod R'$ represents the index that must be shifted to obtain $G'(w)$. The idea here is if we know the composition of each of the granules of $G'$ in terms of granules of $B$ for one period, we can find the composition of any granule of $G'$ beyond a period since the "pattern" repeats itself along the time domain due to the periodic property. The periodic property also ensures that $\vert T_w \vert = \vert T_{w\mod R'} \vert$ since every $w^{\text{th}}$ and $(w+R')^{\text{th}}$ granule of $G'$ will have the same number of granules of $B$. The term $\sum_{w=0}^{k-1}\vert T_{w}\vert$ in Definition \autoref{def:quasicircular} denotes the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$. Since $\vert T_w \vert = \vert T_{w\mod R'} \vert$, the number of granules of $B$ till the $(k-1)^{\text{th}}$ granule of $G'$ becomes $\sum_{w=0}^{k-1}\vert T_{w \mod R'}\vert$ in Definition \autoref{def:quasicircular}.

```{r quasi-circular-example, echo = FALSE, out.width = "100%",fig.cap = "Quasi-circular and aperiodic cyclic granularities illustrated through linear (a) and stack (b) display of time. The linear display shows linear granularities days, weeks, semester weeks, semester Terminology and intuition of quasi-circular granularities are shown. Panel a shows the time domain distributed as linear granularities $B$ (bottom granularity), $G$, $H'$ and quasi-circular granularities $Q(B, G)$ and $Q(B, H')$. Both $Q(B, G)$ and $Q(B, H')$ have repeating patterns unlike $G$ or $H'$ and the patterns are guided by the distribution of $B$ within $G$ and $H'$ and remains constant within a period shown by the hashed arrows. Panel b depicts the intuition through a measuring tape analogy where the folds are not of same length but has same length over a period."}

@@

<!-- Circular granularities illustrated through linear (a and b) and stack (c and d) display of time. The linear display is similar to a measuring tape with the measurement marking distributed as linear granularities day, week and fortnight. The stack display gives the intuition of circular granularities by cutting the measuring tape and stacking on top of one another to represent repetitive pattern of categories. The granules that overlap represent same time domain in the linear display and same category in the stack display. Circular granularities day-of-week ($C_{B, G}$) and day-of-fortnight ($C_{B, H}$) have repeating patterns unlike $B$, $G$ or $H$ which gets manifested more clearly through a stack display of time. -->

@@

```{r aperiodic-example, echo=FALSE, out.width = "100%", fig.cap = "Quasi-circular and aperiodic cyclic granularities illustrated through linear (a) and stack (b) display of time. The linear display shows linear granularities days, weeks, semester weeks, semester week type distributed linearly from past to future. Here a semester lasts for 18.3 weeks, each starting with one week of orientation followed by an in-session period of 6 weeks, a semester-break of 1 week and again an in-session period of 7 weeks after which there is a study break of 1 week before final exams which continue for 1.3 weeks. This pattern remains same for all semester and hence $Q_{H, M}$ with $P = 18.3 weeks$ denoting week of the semester week type will be a quasi-circular granularity with repeating patterns. $A_{B, M}$ will be an aperiodic cyclic granularity since the placement of the semester within an year varies across years (shown through c)."}

knitr::include_graphics("Figs/quasi.png")

```
<!-- #  -->
<!-- # <!-- To consider the exhaustive set of temporal regularities that might exist in the data, we can also define temporal granularities based on the hierarchical structure of a calendar. --> -->
